---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
library(forcats)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
# loading data
tt_data_2020_09_22 <- tt_load('2020-09-22')
expeditions <- tt_data_2020_09_22$expeditions

# filtering 15 most frequent peaks
expeditions_plot_data <- expeditions %>% 
  mutate(peak_name = fct_lump(peak_name, n = 15)) %>% 
  filter(peak_name != "Other")

# aggregating data
expeditions_plot_data <- expeditions_plot_data %>% 
  group_by(peak_name, season) %>%
  summarize(expedition_count = n(), .groups = "drop") %>% 
  ungroup() %>% 
  group_by(peak_name) %>%
  mutate(total_expeditions = sum(expedition_count)) %>% 
  ungroup()

# plotting
expeditions_plot <- expeditions_plot_data %>% 
  ggplot(aes(x = fct_reorder(peak_name, total_expeditions), y = expedition_count, fill = season)) +
  geom_col(position = "stack") +
  coord_flip() +
  scale_fill_viridis_d() +
  theme_light() +
  theme(legend.position = "bottom") +
  # theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "The 15 most popular peaks stacked by season of expedition",
       x = element_blank(),
       y = "Number of expeditions",
       fill = "season")

expeditions_plot
```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
# loading data
tt_data_2019_02_19 <- tt_load('2019-02-19')
phd_data <- tt_data_2019_02_19$phd_by_field

# aggregating data
phd_data_plot <- phd_data %>% 
  group_by(broad_field, year) %>%
  summarize(total_phds = sum(n_phds, na.rm = T), .groups = "drop")

# plotting
phd_plot <- phd_data_plot %>% 
  ggplot(aes(x = year, y = total_phds, color = broad_field)) +
  geom_line(lwd = 1.2) +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  scale_y_continuous(labels = scales::comma_format()) +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  labs(title = "Number of awarded Ph.D.-s in the US by year",
       x = element_blank(),
       y = element_blank(),
       color = "Broad field")

phd_plot
```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
tt_data_2019_11_05 <- tt_load('2019-11-05')
commute_data <- tt_data_2019_11_05$commute

# aggregating data per mode and state (abb)
commute_data_plot <- commute_data %>%
  group_by(state_abb, mode) %>%
  mutate(total_commutes = sum(n, na.rm = T)) %>% 
  select(state_abb, state_region, mode, total_commutes) %>% 
  distinct() %>% 
  pivot_wider(names_from = mode, values_from = total_commutes)

# plotting
commute_plot <- commute_data_plot %>% 
  ggplot(aes(x = Walk, y = Bike, color = state_region)) +
  geom_point(size = 2)  +
  scale_x_log10(labels = scales::comma_format()) +
  scale_y_log10(labels = scales::comma_format(), limits = c(25, NA)) +
  theme_light() +
  labs(title = "Number of people walking vs. biking to work in each USA state",
       x = "Number of ppl walking to work (log N)",
       y = "Number of ppl biking to work (log N)",
       fill = "Region") +
  geom_text(aes(label = state_abb), color = "black")

commute_plot
```
